---
title: "A2_Group5"
author: "Mirre Dona (6156630)"
date: "2023-11-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries
```{r}
library(ISLR) 
library(tidyverse) 
library(readr) # for data import
library(dplyr)
library(magrittr) # for pipes
library(mice) # for missing data
library(ggplot2) # for plotting

```


# Import dataset
```{r}
sleepdata <- read.csv('data/Sleep_health_and_lifestyle_dataset.csv')
View(sleepdata)
```

# Description of the data
The Sleep Health and Lifestyle Dataset, here called sleepdata, consists of 400 rows and 13 columns. The dataset consists of sleep-related data. A lot of variables are presented, including gender, age, occupation, sleep duration, quality of sleep, physical activity levels, stress levels, BMI category, blood pressure, heart rate, daily steps, and, if it is the case, the sleep disorders. This dataset offers the opportunity to gain insights into whether sleep factors and lifestyle factors are related.

To download the dataset, click [here](https://www.kaggle.com/code/jillanisofttech/sleep-health-and-lifestyle-predication-with-94-ac).


# Data cleaning
## Missing data
```{r}
md.pattern(sleepdata)
```
There is no missing data in this dataset!

## Transform variable types
Let's convert the variables to the right type for exploration and prediction later on.
```{r}
#check how many unique values of occupation, to see if it makes sense to convert to factor, or if there are too many unique values
unique(sleepdata$Occupation)
```

```{r}
sleepdata %<>% 
  mutate(Gender   = as.factor(Gender),
         Occupation = as.factor(Occupation),
         BMI.Category = as.factor(BMI.Category)) %>% 
  select(-Person.ID)
```

## Outlier removal
Check outliers

# Exploratory data analysis

```{r}
#summary of the data
summary(sleepdata)

#check for the most important variables

# use ggplot2 to plot the data of sleep quality compared to sleep duration
plot_sleepdata <- sleepdata %>% 
  ggplot(aes(x=Quality.of.Sleep, y=Sleep.Duration)) + 
  geom_point(position = "jitter") + 
  theme_minimal() +
  geom_smooth() +
  labs(title="Sleep quality compared to sleep duration", x="Sleep quality", y="Sleep duration")
plot_sleepdata

# create a boxplot of the sleep quality per sleep duration.
plot_sleepdata2 <- sleepdata %>% mutate(
  Quality.of.Sleep = as.factor(Quality.of.Sleep)
) %>% 
  ggplot(aes(x=Quality.of.Sleep, y=Sleep.Duration)) + 
  geom_boxplot() + 
  theme_minimal() +
  labs(title="Sleep quality compared to sleep duration", x="Sleep quality", y="Sleep duration")
plot_sleepdata2
```
As can be seen, the sleep quality is higher when the sleep duration is higher. This seems logical, because when you sleep longer, you are more rested and therefore feel better. Besides, when the sleep quality is 7, there seems to be the highest variety in sleep duration. It could be tested whether the reason behind this is that most people filled in 7 as sleep quality.

```{r}
# test whether the reason behind the high variety in sleep duration when the sleep quality is 7 is that most people filled in 7 as sleep quality.
sleepdata %>%  
  group_by(Quality.of.Sleep) %>% 
  summarise(n = n())

```
From this, it can be seen that it is not true that most people rated their sleep quality with 7. So other reasons behind this variety can be tested. However, it is clear that less people rated their sleep quality with 4 or 5, which could be a reason why there is less variety in sleep duration when the sleep quality is 4 or 5 (viewed in the boxplot).

```{r}
