---
title: "A2_Group5"
author: "Mirre Dona (6156630), Merel Das (6600034), Lukas Stemerdink (6217265)"
date: "2023-11-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load libraries
```{r}
library(ISLR) 
library(tidyverse) 
library(readr) # for data import
library(dplyr)
library(magrittr) # for pipes
library(mice) # for missing data
library(ggplot2) # for plotting
library(psych)
library(reshape2)

library(randomForest)
library(rpart)
library(rpart.plot)

```

# Import dataset
```{r}
sleepdata <- read.csv('data/Sleep_health_and_lifestyle_dataset.csv')
View(sleepdata)
```

# Description of the data
The Sleep Health and Lifestyle Dataset, here called sleepdata, consists of 400 rows and 13 columns. The dataset consists of sleep-related data. A lot of variables are presented, including gender, age, occupation, sleep duration, quality of sleep, physical activity levels, stress levels, BMI category, blood pressure, heart rate, daily steps, and, if it is the case, the sleep disorders. This dataset offers the opportunity to gain insights into whether sleep factors and lifestyle factors are related.

To download the dataset, click [here](https://www.kaggle.com/code/jillanisofttech/sleep-health-and-lifestyle-predication-with-94-ac).

With this dataset, we aim to predict a person's sleep quality. We chose this response variable as we think it is valuable information for a person. Sleep quality is important for your health and can affect mental-health issues. It also influences how you feel during the day, which affects the activities you undertake and how well you do them.

# Data cleaning
Before we begin, the data should be cleaned for proper use.
## Missing data
```{r}
md.pattern(sleepdata)
```
There is no missing data in this dataset!

## Transform variable types
Let's convert the variables to the right type for exploration and prediction later on.
```{r}
#check how many unique values of occupation, to see if it makes sense to convert to factor, or if there are too many unique values
unique(sleepdata$Occupation)
#The amount of values for the other chr variables were inspected in the R environment
```

```{r}
sleepdata %<>% 
  mutate(Gender   = as.factor(Gender),
         Occupation = as.factor(Occupation),
         BMI.Category = as.factor(BMI.Category),
         Sleep.Disorder = as.factor(Sleep.Disorder))
```
!!nog iets doen met blood pressure

## Select variables
For our analysis and prediction we do not need the person identifier of each case, so we remove this variable.
```{r}
sleepdata %<>% select(-Person.ID)
```

## Outlier removal
We checked for potential outliers based on the IQR criterion. For this, we used boxplot stats per numerical value, and looked how many outliers it produced. 

```{r}
boxplot.stats(sleepdata$Sleep.Duration)$out
boxplot.stats(sleepdata$Quality.of.Sleep)$out
boxplot.stats(sleepdata$Physical.Activity.Level)$out
boxplot.stats(sleepdata$Stress.Level)$out
boxplot.stats(sleepdata$Heart.Rate)$out
boxplot.stats(sleepdata$Daily.Steps)$out

out <- boxplot.stats(sleepdata$Heart.Rate)$out
out_ind <- which(sleepdata$Heart.Rate %in% c(out))
sleepdata[out_ind, ]
```
As can be seen, only the Heart.Rate variable contains outliers; people with a much higher than average heart rate during sleep. When we look at the entries, we see that all these people have a sleep disorder, and that almost all of them are overweight or obese. Because of this, we conclude that these are "expected" outliers, and keep them in the dataset. 

# Exploratory data analysis

To explore the data, we take a look at different factors we have deemed as important: Quality of sleep, sleep duration, and BMI category

```{r}
# create a boxplot of the sleep quality per sleep duration.
plot_sleepdata2 <- sleepdata %>% mutate(
  Quality.of.Sleep = as.factor(Quality.of.Sleep)
) %>% 
  ggplot(aes(x=Quality.of.Sleep, y=Sleep.Duration)) + 
  geom_boxplot(fill = "lightblue") + 
  theme_minimal() +
  labs(title="Sleep quality compared to sleep duration", x="Sleep quality", y="Sleep duration")
plot_sleepdata2

ggplot(sleepdata, aes(x = Gender, y = Quality.of.Sleep)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Sleep Quality by Gender", y = "Quality of Sleep")

sleepdata %>% 
  group_by(Occupation) %>% 
  summarise( 
    median = median(Sleep.Duration),
    min = min(Sleep.Duration),
    max = max(Sleep.Duration),
    mad = mad(Sleep.Duration) # Adding median absolute deviation 
  )

sleepdata %>% 
  ggplot(aes(x = BMI.Category, y = Quality.of.Sleep)) + 
  scale_x_discrete(limits = c("Normal", "Normal Weight", "Overweight", "Obese")) +
  geom_boxplot()

```
From this, it can be seen that it is not true that most people rated their sleep quality with 7. So other reasons behind this variety can be tested. However, it is clear that less people rated their sleep quality with 4 or 5, which could be a reason why there is less variety in sleep duration when the sleep quality is 4 or 5 (viewed in the boxplot).

# Prediction model
As described above, the response variable we chose to predict is quality of sleep.

## Feature selection
To create a well performing model to predict sleep quality, it is important to consider which features to use as predictors. To find the most informative predictors, we will do a correlation analysis and feature importance analysis. 

### Correlation
```{r}
cordata <- sleepdata %>% select(-Gender, -Occupation, -BMI.Category, -Blood.Pressure, -Sleep.Disorder)
#create correlation matrix
corr_mat <- round(cor(cordata),2)

# reduce the size of correlation matrix
melted_corr_mat <- melt(corr_mat)
 
# plotting the correlation heatmap
ggplot(data = melted_corr_mat, aes(x=Var1, y=Var2, fill=value)) + geom_tile()

```

### PCA (of wa anders)

## Simple model selection
### Train, dev, test
```{r}
#train, test, etc.
set.seed(123)
train <- sample(1:nrow(sleepdata), 0.75*nrow(sleepdata)) # 75% of the rows for the training set.
train_data <- sleepdata[train,]
test_data <- sleepdata[-train,] #use the other 25% of the rows for the test set.

#EDA
summary(train_data)

# Train a simple linear regression model
simple_model <- lm(Quality.of.Sleep ~ ., data = train_data)

# Make predictions on the test set
simple_predictions <- test_train %>% 
  mutate(Blood.Pressure = )
  predict(simple_model, newdata = test_data)
```


### Train model

# Classification tree
```{r}
tree_mod <- rpart(Quality.of.Sleep ~ ., data = sleepdata)
rpart.plot(tree_mod) #plot tree
```

# Random forest
```{r}
rf_mod <- randomForest(Quality.of.Sleep ~ ., data = sleepdata)
rf_mod
importance <- importance(rf_mod)

tibble(
  importance = c(importance), 
  variable = rownames(importance)
) %>% 
  ggplot(aes(x = variable, y = importance, fill = variable)) +
  geom_bar(stat = "identity")
```

# Boosting
```{r}
gbm_train <- train(Quality.of.Sleep ~ .,
                   data = sleepdata,
                   method = "gbm",
                   verbose = F, 
                   trControl = cvcontrol) #print bag_train, rf_train, gbm_train models to see accuracy
summary(gbm_train) #var importance plot
predictions <- predict(gbm_train)

```


### Performance evaluation

### Interpretation of model

## Complex model selection





